HO unification from object language to meta language

### 2024-03-19 paper/appendix.tex
appendix in separate file

\documentclass[sigconf,natbib=false,review]{acmart}
\usepackage[]{biblatex}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}
\addbibresource{bib.bib}

\usepackage{myTools}
\usepackage{macros}

% TODO: set this fields
%\setcopyright{cc}
%\setcctype{by}
\copyrightyear{2024}
\acmYear{XXXX 2024}
\acmBooktitle{YYY}
\acmDOI{ZZZZZZZZZZZZ}

% \xspaceaddexceptions{]\}}

\def\elpi{\proglang{elpi}}
\def\coqelpi{\proglang{coq-elpi}}
\def\lambdaprolog{\proglang{$\lambda$-prolog}}
\def\coq{\proglang{coq}}

\newcommand{\library}[1]{\textit{#1}\xspace}
\def\stdpp{\library{stdpp}}
\def\iris{\library{iris}}

\newcommand*{\acronym}[1]{\texttt{#1}\xspace}

\def\ol{\acronym{ol}} % object language
\def\ml{\acronym{ml}} % meta language
\def\lf{\acronym{lf}} % logical framework
\def\ho{\acronym{ho}} % higher order
\def\Forall{$\forall$}

\newcommand{\EqualRel}{\ensuremath{=}}
\newcommand{\nEqualRel}{\ensuremath{\new}}
\newcommand{\UnifRel}{\ensuremath{\simeq}}
\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}

\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}

\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}

\begin{document}

\title{HO unification from object language to meta language}

\author{Enrico Tassi}
\email{enrico.tassi@inria.fr}
\affiliation{%
  \institution{Université Côte d'Azur, Inria}
  % \city{Nice}
  \country{France}}

\author{Davide Fissore}
\email{davide.fissore@inria.fr}
\affiliation{%
  \institution{Université Côte d'Azur, Inria}
  % \city{Nice}
  \country{France}}

\begin{abstract}
  Specifying and implementing a logic from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constuctions~\cite{felty93lics}.

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic DTT.
  Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

  We apply this technique to the implementation of a type-class~\cite{wadler89}
  solver for Coq~\cite{Coq-refman}.
  Type-class solvers are proof search procedures based on
  unification that back-chain designated lemmas, providing essential
  automation to widely used 
  Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
  and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
  constitute our test bed.
\end{abstract}

\keywords{Logic Programming, Meta-Programming, Higher-Order Unification, Proof Automation}

\maketitle

\section{Introduction}
\label{sec:intro}

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873},
Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
Constuctions~\cite{felty93lics}.

The object logic we are interested in is Coq's~\cite{Coq-refman}
Dependent Type Theory (DTT), and we want to code a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman} using the Coq-Elpi~\cite{tassi:hal-01637063}
meta programming framework.
Type-class solvers are unification based proof search procedures
that combine a set of designated lemmas in order to providing essential
automation to widely used Coq libraries.

As the running example we take the \coqIn{Decide} type class,
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class identifies predicates equipped with a decision procedure.
The following three designated lemmas (called \coqIn{Instances} in the
type-class jargon) state that: 1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n} is finite; 2) the predicate \coqIn{nfact n nf},
linking a natural number \coqIn{n} to its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if the
predicate has and if its domain is finite.

\begin{coqcode}
Instance fin_fin n : Finite (fin n).              (* r1 *)
Instance nfact_dec n nf : Decision (nfact n nf).  (* r2 *)
Instance forall_dec A P : Finite A ~$\to$~             (* r3 *)
  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
\end{coqcode}

\noindent Under this context of instances a type-class solver is able to prove
the following statement automatically by back-chaining.

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).       ~\customlabel{goal:g}{(g)}~
\end{coqcode}

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
section~\ref{sec:encodings,sec:lang-spec} and ~\ref{}, is an Higher Order Abstract
Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
constructors:

\begin{elpicode}
type lam  tm -> (tm -> tm) -> tm.     % lambda abstraction
type app  list tm -> tm.              % n-ary application
type all  tm -> (tm -> tm) -> tm.     % forall quantifier
type con  string -> tm.               % constants
\end{elpicode}

\noindent
Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
the concrete syntax to abstract, at the meta level, an expression
\elpiIn{e} over a variable \elpiIn{x}
is <<\elpiIn{x\ e}>>, and square brackets denote a list of
terms separated by comma. As an example we show the encoding of the Coq term
<<\coqIn{~$\forall$~y:t, nfact y 3}>>:

\begin{elpicode}
all (con"t") y\ app[con"nfact", y, con"3"]
\end{elpicode}

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic-programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises;
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

\begin{elpicode}
finite (app[con"fin", N]).                            ~\customlabel{clause:r1}{(r1)}~
decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
  pi w\ decision (app[P, w]).
\end{elpicode}

\noindent
Unfortunately this translation of rule \ref{clause:r3} uses the
predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}.
If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
\ref{goal:g} given below

\begin{elpicode}
decision (all (app[con"fin", con"7"]) y\                 
  app[con"nfact", y, con"3"]).
\end{elpicode}

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths!
%The root cause is that
%\ref{problem:p} is an higher order in DTT, but becomes
%first order in the meta language due to the ``naive'' encoding.

\begin{elpicode}
app[con"nfact", y, con"3"] = app[P, y]                 ~\customlabel{problem:p}{(p)}~
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule \ref{clause:r3} as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
Since \elpiIn{Pm} is an higher-order unification variable
of type \elpiIn{tm -> tm},
with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa}
admits one solution:

\begin{elpicode}
app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
A = app[con"fin", con"7"]              % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal, Elpi runs
the premise <<\elpiIn{link Pm A P}>> that is in charge of bringing the
assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:

\begin{elpicode}
P = lam A a\ app[con"nfact", a, con"3"]
\end{elpicode}

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.
In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}):

\begin{elpicode}
decision (app[lam A (a\ app[con"nfact", a, con"3"]), w])
\end{elpicode}

\noindent
In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
the following unification problem has no solution:

\begin{elpicode}
app[lam A (a\ app[con"nfact", a, con"3"]), x] =
app[con"nfact", N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in the running example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.

\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment $\llambda$~\cite{miller92jsc}.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi that we want to use to
implement a form of proof automation is strikingly similar, since it
it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher-order problems in
$\llambda$.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example

\begin{tabular}{lcl}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"}\\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"} \\
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}

\noindent
One could ignore this similarity, and ``just'' describe the object language
unification procedure in the meta language, that is crafting a \elpiIn{unif}
predicate to be used as follows in rule \ref{clause:r3}:

\begin{elpicode}
decision X :- unif X (all A x\ app[P, x]), finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
This choice would underuse the logic programming engine provided by
the metalanguage since by removing any datum from the head of rules
indexing degenerates. Moreover the unification procedure built in the
meta language is likely to be faster than one implemented in it,
especially if the meta language is interpreted as Elpi is.

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
TODO extend \Eo and \Ee with reflexivity on uvars.

\newcommand{\specunif}[3]{
  #3_i \in \llambda \Rightarrow %
    \exists \rho, %
      \rho #3_1 #1 \rho #3_2  %
        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho 
}

\newcommand{\unifcorrect}[3]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow 
          \rho #3_1 #1 \rho #3_2  %
}

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho 
}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
and we assume that the unification of our meta language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress XXX improve....

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \solve{} on the set of links. We claim the following:

\begin{proposition}[Simulation]\label{prop:sumul}
$\forall \mathcal{S}, \forall \mathcal{N}$
$$
  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
$$
\end{proposition}

\noindent
That is, the two executions give the same result. Moreover:

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$

\begin{proposition}[Properties of \Uo{}]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
% \forall \rho'\rho, 
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \llambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
\end{gather}
\end{proposition}

Properties \ref{prop:correct} and \ref{prop:complete} state, respectively, that
in \llambda the implementation of \Uo is correct, complete and returns the most
general unifier.

Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

\begin{elpicode}
app [F, con"a"] = app[con"f", con"a", con"a"]   ~\customlabel{problem:q}{(q)}~
F = lam x\ app[con"f",x,x]                      ~\customlabel{heuristic:h}{(h)}~
\end{elpicode}

\noindent
Instead of rejecting it our scheme accepts it and guarantees that if
\ref{heuristic:h} is given (after the compilation part of the scheme, as
a run time hint) then ...

\subsection{The intuition in a nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a unification variable $h$
and an accessory link, that represent a suspended unification problem
$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
We now define ``problematic'' formally.

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F~y~x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
makes $\rho t = \lambda x.\lambda y.f x y$
that is the eta long form of $f$.

\begin{definition}[\maybebeta]
  $\maybebeta = \{ X t_1 \ldots t_n | t \not\in \llambda \}$.
\end{definition}

\noindent
An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
that makes the resulting term fall outside of \maybebeta.

\begin{definition}[Subterm $\mathcal{P}(t)$]
$$
\begin{array}{l}
t \in \mathcal{P}(t)\\
t = f t_1\ldots t_n \Rightarrow \mathcal{P}(t_i) \subseteq \mathcal{P}(t) \land f \in \mathcal{P}(t)\\
t = \lambda x.t' \Rightarrow \mathcal{P}(t') \subseteq \mathcal{P}(t)\\
\end{array}
$$
\end{definition}

\noindent
We write $\mathcal{P}(X) = \bigcup_{t\in X} \mathcal{P}(t)$.

\marginpar{TODO: find better word}
\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
\begin{definition}[Normal form]
Given a set of terms $X \subseteq \Ho{}$,
$$
\normal(X) =
\mathcal{P}(X) \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
$$
\end{definition}

\noindent
We write $\sigma X = \{ \sigma t | t \in X\}$.

\begin{proposition}[Normal form preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
$$
\normal(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
\Rightarrow \normal((\sigma \cup \sigma')\mathcal{T})
$$
\end{proposition}

\noindent
In particular this guarantees that is we start from normal terms
we never introduce eta-long or non-beta-normal terms in $\sigma'$.

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta.

\section{Alternative encodings and related work}

Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.

Our encoding of DTT may look ``semi shallow'' since we use the meta-language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

\begin{elpicode}
finite (fin N).
decision (nfact N NF).
decision (all A x\ P x) :- finite A, pi x\ decision (P x).
\end{elpicode}

\noindent
There are two reasons for dismissing this encoding. The first one is that
in DTT it is not always possible to adopt it since the type system
of the meta language is too weak to accommodate terms with a variable arity,
like the following example:

\begin{coqcode}
Fixpoint arr T n := if n is S m then T -> arr T m else T.
Definition sum n : arr nat n := ...
Check sum 2   7 8   : nat.
Check sum 3   7 8 9 : nat.
\end{coqcode}

\noindent
The second reason is the encoding for Coq is used for meta programming the
system, hence it must accommodate the manipulation of terms that are now
know in advance (not even defined in Coq) without using introspection
primitives such as Prologs's \texttt{functor} and \texttt{arg}.

In the literature we could find a few related encoding of DTT.
TODO In~\cite{felty93lics} is related and make the
discrepancy between the types of ML and DTT visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

TODO This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.

None of the encodings above provide a solution to our problem.

\section{Preliminaries: \Fo{} and \Ho}
\label{sec:lang-spec}

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax. We keep these languages
minimal, for example, we omit the \elpiIn{all} quantifier of DTT we used
in the example in Section~\ref{sec:intro} together with the type notation of
terms carried by the \elpiIn{lam} constructor. 
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{\Fo{} and \Ho language}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
the arguments of an higher order unification variable are via the \elpiIn{fapp}
constructor. For example in the statement of the instance \coqIn{forall_dec} the
term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is
a memory address and \elpiIn{x} is a bound variable.

In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
We say that the unification variable \elpiIn{uva N L} is in \llambda iff
\elpiIn{distinct L} holds.

\begin{elpicode}
type distinct list A -> o.
distinct [].
distinct [X|XS] :- name X, not(mem X XS), distinct XS.
\end{elpicode}

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.
\footnote{one could always load name x for every x under a pi and get rid of the name builtin}
The
compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
reasons, so we don't assume all out terms are in \llambda but rather test. \textbf{\textcolor{red}{what?? }}

In both languages unification variables are identified by a natural number,
which can be seen as a memory address. The memory and its associated
operations are described below:

\begin{elpicode}
typeabbrev memory A (list (option A)).
type set?   nat -> memory A -> A -> o.
type unset? nat -> memory A -> o.
type assign nat -> memory A -> A -> memory A -> o.
\end{elpicode}

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value.

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it in order to enable the instantiation of a single
solution to different scopes. On the contrary solutions to \Fo variables
are plain terms.

\begin{elpicode}
  typeabbrev fsubst (memory ftm).
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst (memory assmt).
\end{elpicode}

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO explain.

\subsection{Equational theory and Unification}

here we give the functions/signatures to express the properties 3-5 in the problem
statement

Together with the description of the terms of the language, we need some
auxiliary functions to perform operations like term equality, unification,
dereferencing. There predicates are supposed to be implemented in the OL and the
ML following respectively their specification. In the following few paragraphs
we give a brief proposition of them.

\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}

Since in our encoding we explicitly carry a substitution we need to
define the operation that applies it to a term. Its proposition
takes care to normalize (flatten) applications, for example
it turns
\elpiIn{(app [app [con"f",con"a"],con"b"])}
into
\elpiIn{(app [con"f",con"a",con"b"])}.

dereference variables of the two languages. This is particularly useful to check
if two terms in the OL (resp. in the ML) are equal. The constructors
representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
make the dereferencing procedure to recurse over the their subterms. The code
below display this behavior for the dereferencing performed by the OL:

\begin{elpicode}
type fder fsubst -> fm -> fm -> o.
fder S (fapp A) (fapp B) :- map (fder S) A B.
fder S (flam F) (flam G) :- 
  pi x\ fder S x x => fder S (F x) (G x).
fder _ (fcon C) (fcon C).
fder S (fuva N) R :- set? N S T, fder S T R.
fder S (fuva N) (fuva N) :- unset? N S.

type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fder}{\rho s}$)~
fderef S T R :- fder S T T', napp T' R.

type napp fm -> fm -> o.
\end{elpicode}

\noindent
TODO explain napp.
We use the cut operator to keep the code compact. It is possible to rewrite
the rule for application traversal so that it is mutually exclusive with the
first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.

The corresponding code for \Ho is similar, we only show the last two rules that
differ in a substantial way:

\begin{elpicode}
type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
deref S (app A) (app B) :- map (deref S) A B.
deref S (lam F) (lam G) :- 
  pi x\ deref S x x => deref S (F x) (G x).
deref _ (con C) (con C).
deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
type move assignment -> list tm -> tm -> o.
move (abs Bo)        [H|L] R :- move (Bo H) L R.
move (val A)         []    A :- !.
move (val (uva N A)) L     (uva N X) :- std.append A L X.
\end{elpicode}

\noindent
TODO: no need to napp, see the beta section.
Note that when the substitution \elpiIn{S} maps a unification variable
\elpiIn{N} to an assignment \elpiIn{F} we ....

>>>>

\section*{Appendix}
